### 2018-03-08 paper/sexSelectionSpacing-ver1.tex
Explain result table structure and measures

The main downside of this set-up is that interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes more likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.
It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

@@

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
Finally, I include standard survival curves, which shows the ``weight''
that each period has, or, in other words, how many births take place in
each period.

@@

Results are presented as the estimated percent boys born by length of birth spacing
using graphs.
For each graph, the extent to which the percent boys is statistically 
significantly above the natural sex ratio indicates the use of sex selection.

[Moved from estimation strategy]
In addition to information about spacing, the survival curves also provide 
``weighting'' for the associated percentage boys born.
The steeper the survival curve, the more weight should be assigned to a given
spell period because it is based on more births, 
whereas a period with a flat survival curve should be given little weight because the 
percentage boys is based on few births.

[Difference in approach: use the sex composition of prior
children, rather the sex of the last child.
Problem is that the cell sizes become too small if we further split up
the higher spell cells.

Define cell]

The main question of interest here is how the duration between one
birth and the next is related to son preference.
The hypothese examined here is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
used.
We would still expect parents to \emph{begin} trying to conceive earlier
after the birth of a girl than after a boy, but the abortions will automatically
increase the length of the space, and may even lead to longer spacing after the
birth of girls than after boys.

To provide a first look at how the hypothese holds up, Tables 
\ref{tab:median_sex_ratio_low},
\ref{tab:median_sex_ratio_med}, and
\ref{tab:median_sex_ratio_high}
show predicted median duration and sex ratios for the three different
education levels by spell, sex composition, and periods.
These predictions are based on the estimation results employing the method 
detailed in Section \ref{sec:strategy} and are calculated using the means of 
the continuous explanatory variables and the majority category for categorical 
explanatory variables.
The median duration is the number of months it is predicted to take for
half of the women who are predicted to have a birth by the end of spell length
covered.
For example, if 80\% of women with a given set of characteristics are predicted to
have a birth by the end of the end of spell, the median duration is the number of
months it would take for 40\% of women to have a birth.
The sex ratio is the percent of births that result in a son, based on all the 
births predicted to happen over the spell.
If the estimations are based on 100 or fewer births in a cell the predictions
for that cell are not presented.
Furthermore, those cells with 500 or fewer births are marked.

@@

\subsection{[Specific spell/education/area combinations of special interest]}
[Only a subset of the planned analyses to demonstrate the method and interpretation]

>>>>

The main downside of this set-up is that interpretation of the estimated 
coefficients is challenging;
the coefficients show the change in hazards relative to the base outcome, 
here no birth, rather than the hazard of an event.
Hence, a positive coefficient does not necessarily imply that the associated 
exit state becomes more likely as the explanatory variable increases because the 
probability of the other exit state may increase even more \citep{thomas96}.

It is, however, straightforward to calculate the predicted probabilities of 
having a boy and of having a girl for each $t$ within a spell, conditional on 
a set of explanatory variables and not having had a child before that period.
The predicted probability of having a boy in period $t$ for a given set of 
explanatory variable values, $\mathbf{Z}_k$ and $\mathbf{X}_k$, is
\begin{equation}
P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt}, t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{1t}' \mathbf{Z}_{kt} + \beta_1' \mathbf{X}_{k} )}
{1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt} ' \mathbf{Z}_{kt} + \beta_l ' \mathbf{X}_{k})},
\label{eq:probability_boy}
\end{equation}
and the predicted probability of having a girl is
\begin{equation}
P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t ) 
=  
\frac{ \exp(D_j(t) + \alpha_{2t}'\mathbf{Z}_{kt} + \beta_2'\mathbf{X}_{k} )}
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{kt} + \beta_l'\mathbf{X}_{k})}.
\label{eq:probability_girl}
\end{equation}
Within each period the probability of not having a birth in period $t$
is $1-P(b_{t})-P(g_{t})$.

@@

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.

@@

% [Hypotheses / questions]

The overall hypothesis is that spacing patterns significantly changed 
as prenatal sex determination became available.
To address this, I examine three questions in order of importance.

First, did the spacing after girls increase relative to the spacing
after boys?
The underlying hypothese is that in the absence of sex selection son 
preference leads to shorter spacing after the birth of a girl than after the 
birth of a boy, whereas son preference increases the spacing after the birth 
of a girl relative to after the birth of a boy when prenatal sex selection is 
used.
We would still expect parents to \emph{try} to conceive earlier after the birth 
of a girl than after a boy, but the abortions will increase the 
length of the space, and may even lead to longer spacing after the
birth of girls than after boys.
[TK something on general changes in spacing? Or hold off until discussing the
results]

Second, do the shapes of survival curves differ across sex compositions 
within a spell?
The shapes of the survival curves provides clues to whether parents are
still trying to conceive earlier after girls than boys.
I expect this to be the case if there is still a strong preference for
boys, and, as discussed above, this behavior may even become more pronounced
with access to sex selection.
At a more technical level, the shapes of the survival curve serve as an
indicator for the presence of non-proportionality.

Finally, do parents change their decision on sex selection within a spell?
There are two reasons why this is interesting.
If parents change their mind the standard method of observing sex ratios 
will not capture the full extent of the use of sex selection.
Furthermore, it tells us something about parents' preferences for ideal
spacing between births and how strong their son preference is.

% [How to read tables]
Tables \ref{tab:median_sex_ratio_low}, \ref{tab:median_sex_ratio_med}, and
\ref{tab:median_sex_ratio_high}
show predicted median duration and sex ratios by period, spell, and sex 
composition for the three different education levels separated by area of residence. 
The predictions are based on the estimation method described in 
Section \ref{sec:strategy}.
The standard errors for both measures are based on bootstrapping, where the 
original sample is resampled with replacement and the method is applied to 
the new sample.

The predicted median duration shown is the weighted average over the women in the 
sample's predicted median durations, using the individual probabilities of a birth 
as weights.
For each woman in a sample, I calculate her probability of ending the spell 
with a birth, and her contribution to median spacing is how long it will take
from the beginning of the spell to where she is predicted to reach half of that probability.
For example, if a woman has a 80\% probability of having a birth by the end of 
the spell, her median duration is the number of months it takes to reach
40\% probability of having exited with a birth (equivalent to the 60\% point
in a standard survival curve because that captures how many have not yet
exited).

For second through fourth spell I show whether durations for sex composition
other than only girls are statistically significantly different from the
duration with only girls.
The statistical significance is based on the bootstrapped differences in
durations across sex compositions.
For each new sample the difference in duration is calculated and these
differences are used to calculate the standard error and the level of
significance for the difference in the original sample.

The cleanest test is comparing durations after only boys with durations after
only girls.
However, because of falling fertility the number of births to women with 
only sons becomes small in the later periods, especially for the third and 
fourth spell.
Hence, it is possible to have substantial differences in spacing that are
not statistically significant because of low power.
Furthermore, it is, in principle, possible to estimate the model taking 
into account the ordering of children.
This would, however, further lower the power of the test, by leading to 
one extra group for the third spell and four extra groups for the fourth spell.

The predicted sex ratio is the weighted average over the women in the sample's 
individual predicted sex ratios, using the individual probabilities of a birth by 
the end of the spell as weights.
Each woman's predicted sex ratio is the weighted average of the predicted
percentage boys over the months in the spell calculated using equation 
(\ref{eq:probability_son}) and the probability of giving birth in each month 
as weights.%
\footnote{
Imagine a spell has two periods and that the estimated percentage boys for a
woman are 54\% and 66\% and that the likelihood of exiting the spell is 20\% and 40\%
(in other words, there is 40\% chance that she does not have a birth by the end of 
the spell).
This woman' percentage boys is then $\frac{54*0.2+66*0.4}{0.2+0.4} = 62$.
}
The predicted sex ratio captures the percent boys that will be born to women 
in the sample once child bearing for that spell is over.

Each cell's predicted precent boys is tested against the natural percentage 
boys using the bootstrapped standard errors.
The natural sex ratio is approximately 105 boys to 100 girls or
51.2\% \citep{ben-porath76b,jacobsen99,Portner2015b}.
It is possible for the predicted percentage boys to be different from 
the natural rate either because of natural variation, any remaining
recall error, or sex selection.

@@

\subsection{Differences in Survival Curves Across Sex Compositions}

\subsection{Does Sex Selection Change within Spells?}

\section{Discussion}
